Implementation of a minimum branch-decomposition and branch-width algorithm for planar graphs.

### 2024-08-19 paper/images/carving.tex
Update paper

		% Draw the edges between nodes with labels
		\draw[line width=0.5mm] (a1) -- node[left] {1} (a2);
		\draw[line width=0.5mm] (a2) -- node[left] {2} (a3);
		\draw[line width=0.5mm] (a2) -- node[above] {3} (a4);
		\draw[line width=0.5mm] (a4) -- node[right] {4} (a5);
		\draw[line width=0.5mm] (a3) -- node[above] {5} (a5);
	\end{scope}

	\begin{scope}[xshift=6cm, yshift=0cm]
		\node (G) at (1, 6) {Branch Decomposition of $G$};

		\node[circle, draw, color=black, fill=white] (b1) at (1, 4) {(1,2)};
		\node[circle, draw, color=black, fill=white] (b3) at (0, 3) {(2,4)};
		\node[circle, draw, color=black, fill=white] (i1) at (1, 3) {};
		\node[circle, draw, color=black, fill=white] (i2) at (1.5, 2) {};
		\node[circle, draw, color=black, fill=white] (i3) at (1, 1) {};
		\node[circle, draw, color=black, fill=white] (b2) at (2.5, 2) {(2,3)};
		\node[circle, draw, color=black, fill=white] (b4) at (0, 1) {(4,5)};
		\node[circle, draw, color=black, fill=white] (b5) at (1, 0) {(3,5)};

		% Draw the edges between nodes with labels
		\draw[line width=0.5mm, color=black] (b1) -- (i1);
		\draw[line width=0.5mm, color=black] (b3) -- (i1);
		\draw[line width=0.5mm, color=black] (i1) -- (i2);
		\draw[line width=0.5mm, color=black] (b2) -- (i2);
		\draw[line width=0.5mm, color=black] (i2) -- (i3);
		\draw[line width=0.5mm, color=black] (i3) -- (b4);
		\draw[line width=0.5mm, color=black] (i3) -- (b5);
	\end{scope}
\end{tikzpicture}

>>>>

		% Draw the edges between nodes with labels
		\draw[line width=0.5mm] (a1) -- node[left] {1} (a2);
		\draw[line width=0.5mm] (a2) -- node[left] {2} (a3);
		\draw[line width=0.5mm] (a2) -- node[above] {3} (a4);
		\draw[line width=0.5mm] (a4) -- node[right] {4} (a5);
		\draw[line width=0.5mm] (a3) -- node[above] {5} (a5);

		\node[opacity=0] at (-1, 3.5) (start) {};
		\node[opacity=0] at (-1, 0.5) (end) {};
		\draw[dotted, thick] (start) .. controls +(2,0) and +(2,0) .. (end);
	\end{scope}

	\begin{scope}[xshift=6cm, yshift=0cm]
		\node (G) at (1, 6) {Carving Decomposition of $G$};
		\node (G) at (1, 5.5) {of width 3};

		\node[circle, draw, color=black, fill=white] (b1) at (1, 4) {1};
		\node[circle, draw, color=black, fill=white] (b3) at (0, 3) {2};
		\node[circle, draw, color=black, fill=black] (i1) at (1, 3) {};
		\node[circle, draw, color=black, fill=black] (i2) at (1.5, 2) {};
		\node[circle, draw, color=black, fill=black] (i3) at (1, 1) {};
		\node[circle, draw, color=black, fill=white] (b2) at (2.5, 2) {3};
		\node[circle, draw, color=black, fill=white] (b4) at (0, 1) {4};
		\node[circle, draw, color=black, fill=white] (b5) at (1, 0) {5};

		% Draw the edges between nodes with labels
		\draw[line width=0.5mm, color=black] (b1) -- (i1);
		\draw[line width=0.5mm, color=black] (b3) -- (i1);
		\draw[line width=0.5mm, color=black] (i1) -- (i2);
		\draw[line width=0.5mm, color=black] (b2) -- (i2);
		\draw[line width=0.5mm, color=black] (i2) -- (i3);
		\draw[line width=0.5mm, color=black] (i3) -- (b4);
		\draw[line width=0.5mm, color=black] (i3) -- (b5);

		\node (G) at (-0.5, 4.25) {\{1,2,3\}};

		\node[opacity=0] at (0,4) (start) {};
		\node[opacity=0] at (0,2) (end) {};
		\draw[dotted, thick] (start) to[bend left=60] (end);
	\end{scope}

	\begin{scope}[xshift=12cm, yshift=0cm]
		\node (G) at (1, 6) {Minimum Carving Decomposition of $G$};
		\node (G) at (1, 5.5) { of width 3};

		\node[circle, draw, color=black, fill=white] (b1) at (1, 4) {1};
		\node[circle, draw, color=black, fill=white] (b3) at (0, 3) {2};
		\node[circle, draw, color=black, fill=black] (i1) at (1, 3) {};
		\node[circle, draw, color=black, fill=black] (i2) at (1.5, 2) {};
		\node[circle, draw, color=black, fill=black] (i3) at (1, 1) {};
		\node[circle, draw, color=black, fill=white] (b2) at (2.5, 2) {3};
		\node[circle, draw, color=black, fill=white] (b4) at (0, 1) {4};
		\node[circle, draw, color=black, fill=white] (b5) at (1, 0) {5};

		% Draw the edges between nodes with labels
		\draw[line width=0.5mm, color=black] (b1) -- (i1);
		\draw[line width=0.5mm, color=black] (b3) -- (i1);
		\draw[line width=0.5mm, color=black] (i1) -- (i2);
		\draw[line width=0.5mm, color=black] (b2) -- (i2);
		\draw[line width=0.5mm, color=black] (i2) -- (i3);
		\draw[line width=0.5mm, color=black] (i3) -- (b4);
		\draw[line width=0.5mm, color=black] (i3) -- (b5);

		\node (G) at (-0.5, 4.25) {\{1,2,3\}};

		\node[opacity=0] at (0,4) (start) {};
		\node[opacity=0] at (0,2) (end) {};
		\draw[dotted, thick] (start) to[bend left=60] (end);
	\end{scope}
\end{tikzpicture}

### 2024-08-19 paper/main.tex
Fix code snippets line ranges
Update paper

\title{Implementation of a minimal branch-decomposition algorithm for planar graphs.}
\author{Kristoffer Højelse}
\date{February 2024}

@@

\section{Introduction}
	Some graph optimization problems can be solved efficiently for graphs of small branchwidth.\cite{CNP+11}

@@

	\todo{which? counting Hamiltonian cycles of planar cubic graphs}

@@

@@

	\textbf{Note.} Other authors might instead call this definition an undirected labelled pseudograph, with edges having own identity.

@@

	A \textit{multi-graph} is a graph that is loop-less.

	\begin{center}
		\input{images/multigraph.tex}
	\end{center}

	A multi-graph $G$ is \textit{simple}, if it has no parallel edges; in other words, if elements of $E(G)$ are pair-wise distinct.

@@

		A contraction is a function that given a multi-graph $G$ and pair of distinct vertices $u,v \in V(G)$ such that $\{u,v\} \in E(G)$, then for all edges $e \in \mathbb{E}(G)$ if $\phi_G(e) = \{u,v\}$ then removes $e$ else if $\phi_G(e) = \{v,w\}$ then $\phi_G(e) = \{u,w\}$, and finally returns the resulting graph.
	\end{definition}

@@

	An \textit{internal vertex} $v$ of a tree $T$, is a vertex $v \in V(T) - L(T)$ that is not a leaf. An internal vertex therefore has degree at least 2.

@@

@@

	Removing any edge $e \in \mathbb{E}(B_G)$ partitions $B_G$ into 2 trees $P_e$ and $Q_e$. The set $L(P_e) \cap L(Q_e)$ is called a \textit{middle set} of $B_G$ given $e$, denoted $Z(B_G,e)$. The maximal cardinality of any middle set of $B_G$ given any $e$ of $B_G$ is the width of $B_G$; in other words, the width of $B_G$ is $\max\{ |Z(B, e)| \st e \in \mathbb{E}(B) \}$.

	There might exist many branch decompositions of a graph $G$.

	A \textit{Minimal Branch Decomposition} of $G$ is any branch decomposition of $G$ of minimal width among all branch decompositions of $G$.

@@

	Removing any edge $e \in \mathbb{E}(C_G)$ partitions $C_G$ into 2 trees $P_e$ and $Q_e$. The set $L(P_e) \cap L(Q_e)$ is called a \textit{middle set} of $C_G$ given $e$, denoted $Z(C_G,e)$. The maximal cardinality of any middle set of $C_G$ given any $e$ of $C_G$ is the width of $C_G$; in other words, the width of $C_G$ is $\max\{ |Z(C_G, e)| \st e \in \mathbb{E}(C_G) \}$.

	There might exist many carving decompositions of a graph $G$.

	A \textit{Minimal Carving Decomposition} of $G$ is any carving decomposition of $G$ of minimal width among all carving decompositions of $G$.

@@

	Given a simple connected planar graph $G$, the algorithm computes a minimal branch decomposition of $G$.

	To compute a minimal branch decomposition of $G$, the algorithm first computes the medial graph $G^\times$ of $G$, secondly computes a minimal carving decomposition of $G^\times$, and finally computes a minimal branch decomposition of $G$ from the minimal carving decomposition of $G^\times$.

	To compute the minimal carving decomposition of $G^\times$, the algorithm uses a contraction algorithm that finds a series of contractions that does not increase the carving width and then assembles the series of contractions into a minimal carving decomposition of $G^\times$.

@@

	The main computational problem of this paper is \textsc{The Planar Minimal Branch Decomposition Problem}.

	\begin{definition}\label{def:mbdp}
		\textsc{The Planar Minimal Branch Decomposition Problem}

@@

		Output: A minimal branch decomposition of $G$.
	\end{definition}

	The algorithm described in this paper solves \textsc{The Planar Minimal Branch Decomposition Problem} in polynomial time.

@@

	Considering a plane graph $G$, you can compute a minimal branch decomposition $(B_G, \delta_G)$ of $G$ from a minimal carving decomposition $(C_{G^\times}, \lambda_{G^\times})$ of the medial graph $G^\times$ of $G$.

@@

	\begin{problem}\label{problem:medial}
		Given a plane graph $G$, output a medial graph $G^\times$, along with a bijectional relation between medial nodes $V(G^\times)$ and vertex-pairs $E(G)$.
	\end{problem}

	\begin{problem}\label{problem:mcd}
		Given a plane graph $M$, output a minimal carving decomposition of $M$.
	\end{problem}

	\begin{problem}\label{problem:mcd-to-mbd}
		Given a minimal carving decomposition of a medial graph of $G$, output a minimal branch decomposition of $G$.
	\end{problem}

@@

	\pyfile{branch_decomposition.py}{0}{100000}

@@

		\pyfile{medial_graph.py}{4}{24}

	\subsection{Minimal Branch Decomposition from Minimal Carving Decomposition}

@@

	\subsection{Minimal Carving Decomposition}

@@

		By doing a series of edge contractions on a graph $M$, where the carving width does not increase until 3 vertices remain, then the series of contracted edges along with the three vertices, can be assembled into a minimal carving decomposition of $M$.

		The implementation finds a nonincreasing contraction by doing a linear search over every edge. No consideration has yet been given to any potential clever orderings of the edges that might improve the running time.

@@

		\pyfile{carving_decomposition.py}{8}{48}

@@

			\pyfile{carving_width.py}{0}{8}
			\pyfile{carving_width.py}{132}{157}

@@

		\pyfile{carving_width.py}{7}{9}
		\pyfile{carving_width.py}{78}{130}

		\subsubsection{Quiet subgraph}

@@

		\subsubsection{Dual graph}\label{impl:dual}

@@

			The dual has a vertex for each face of the input graph. The faces are found by selecting an unmarked half-edge, and then marking all the edges of the face it belongs to.

@@

			\pyfile{dual_graph.py}{21}{22}

			\pyfile{dual_graph.py}{4}{50}

@@

>>>>

\title{Implementation of a minimum branch-decomposition and branch-width algorithm for planar graphs.}
\author{Kristoffer Højelse}
\date{February 2024}

@@

\section{Introduction}

	\todo{change graph to multigraph}

	In complexity theory, an NP-hard problem has no known algorithm with all 3 of these properties: (1) runs in polynomial time, (2) gives an exact solution, and (3) runs on all instances of the problem.

	You can find brute-force algorithms for NP-hard problems that both give an exact solution and run on all instances of the problem, but they run in exponential time.

	You can find approximation algorithms for NP-hard problems that both run in polynomial time and run on all instances, but they give an approximate solution.

	Finally, you can find parameterized algorithms for NP-hard problems that both run in polynomial time and give an exact solution, but they only run on instances of the problem where some parameter is small. Treewidth is a common such parameter.

	Treewidth is a measure of how tree-like a graph is, a value of 1 is exactly a tree and larger values are less tree-like. For some NP-hard problems, if a graph is sufficiently tree-like, then a parameterized algorithm can find an exact solution in polynomial time. For instance, the best-known running time for the \textsc{The Maximum Independent Set Problem} is exponential in $n$, particularly $O(1.1996^n \cdot n^{O(1)})$\cite{XN17}, but can be solved in $O(2^k \cdot k \cdot n)$ time for graphs of treewidth at most $k$, which is linear in $n$ and exponential in $k$.

	Branch width and carving width are related to treewidth, and some NP-hard problems can be solved efficiently for graphs of small branchwidth.\cite{CNP+11}

	Tree-, branch- and carving-width are all related. The treewidth of a graph is at least bw-1 and at most floor(3/2bw)-1.\cite{RS91}

	The carving width is at least half the branch width and is at most the degree times the branch width.\cite{ED18}

	You might have a parameterized algorithm with tree- or branch-width as a parameter, but to determine if a graph has a small tree- or branch-width, you might need to compute the tree- or branch-decomposition first. There is no known optimal tree decomposition/treewidth algorithm even for planar graphs.

	A planar graph is, informally, a graph that you can draw on a plane surface without any edges crossing. There might be multiple such drawings of a planar graph. A particular such drawing is called a plane graph.

	The algorithm by Seymour and Thomas\cite{ST93} computes a minimum branch-decomposition of a planar graph in $O(n^4)$ time.

	You can obtain an approximate optimum tree decomposition from the branch decomposition. So computing an approximate treewidth for planar graphs is possible in polynomial time.

	The concepts of medial graph and dual graph are used in the algorithm. Both are planar graphs, that are constructed from a plane graph.

	The dual graph has vertices for faces and edges connecting the faces that share an edge. You might imagine a map of countries, with edges for borders, and a vertex for each country, plus a vertex for the surrounding sea. You could for instance query such a network for which countries are neighbors, or which countries are landlocked.

	The medial graph also considers the faces but has vertices for edges, that are connected if the original edges were consecutive in some face. I have found no good metaphors for this.

@@

	\todo{Read params book about tree width}

@@

	This section defines the terms and concepts used in the algorithm.

	The algorithm deals with graphs that might have parallel edges. Distinguishing between two parallel edges is crucial. Therefore, the implementation gives labels to both vertices and edges, by encoding a graph as an adjacency list of edge IDs and a map from unique edge IDs to its vertex-pair.

@@

	\textbf{Note.} Normally \textit{graph} refers to a simple graph, usually also without labels, which are normally encoded by an adjacency list or matrix. In this paper, a graph is a multigraph with labels.

@@

	A graph $G$ is \textit{simple}, if it has no parallel edges; in other words, if all elements of $E(G)$ are pair-wise distinct.

@@

		A contraction is a function that given a multi-graph (with no self-loops) $G$ and pair of distinct vertices $u,v$ of some edge $\{u,v\} \in E(G)$, removes $e$ if $\phi_G(e) = \{u,v\}$ and makes any edge that is adjacent to $v$, $\phi_G(e) = \{w,v\}$, adjacent to $u$ instead, $\phi_G(e) = \{w,u\}$, and finally returns the resulting graph.
	\end{definition}

	\begin{corollary}
		The resulting graph of a contraction will not have any self-loops.
	\end{corollary}

@@

	An \textit{internal vertex} $v$ of a tree $T$, is a vertex $v \in V(T) - L(T)$ that is not a leaf. An internal vertex therefore has at least degree 2.

@@

	Removing any edge $e \in \mathbb{E}(B_G)$ partitions $B_G$ into 2 trees $P_e$ and $Q_e$. The set $\cup\{\phi_G(e) \st e \in L(P_e)\} \cap \cup\{\phi_G(e) \st e \in L(Q_e)\}$ is called a \textit{middle set} of $B_G$ given $e$, denoted $Z(B_G,e)$. The maximal cardinality of any middle set of $B_G$ given any $e \in \mathbb{E}(B_G)$ is the width of $B_G$; in other words, the width of $B_G$ is $\max\{ |Z(B, e)| \st e \in \mathbb{E}(B) \}$.

	\todo{middle set definition is convoluted}

	A \textit{Minimum Branch Decomposition} of $G$ is any branch decomposition of $G$ of minimum width among all branch decompositions of $G$, as there might exist many branch decompositions of a graph $G$.

@@

@@

	Removing any edge $e \in \mathbb{E}(C_G)$ partitions $C_G$ into 2 trees $P_e$ and $Q_e$. The set of edges $e \in \mathbb{E}(G)$ that cross $L(P_e)$ and $L(Q_e)$ is called a \textit{crossing set} of $C_G$ given an $e \in \mathbb{E}(C_G)$, denoted $Y(C_G,e)$. The maximum cardinality of any crossing set of $C_G$ given any $e \in \mathbb{E}(C_G)$ is the width of $C_G$; in other words, the width of $C_G$ is $\max\{ |Y(C, e)| \st e \in \mathbb{E}(C) \}$.

	A \textit{Minimum Carving Decomposition} of $G$ is any carving decomposition of $G$ of minimum width among all carving decompositions of $G$, as there might exist many carving decompositions of a graph $G$.

	\begin{center}
		\input{images/carving.tex}
	\end{center}

	\todo{find example with non-mimimum carving decomposition}

@@

	Given a simple connected planar graph $G$, the algorithm computes a minimum branch decomposition of $G$.

	To compute a minimum branch decomposition of $G$, the algorithm first computes the medial graph $G^\times$ of $G$, secondly computes a minimum carving decomposition of $G^\times$, and finally computes a minimum branch decomposition of $G$ from the minimum carving decomposition of $G^\times$.

	To compute the minimum carving decomposition of $G^\times$, the algorithm uses a contraction algorithm that finds a series of contractions that does not increase the carving width and then assembles the series of contractions into a minimum carving decomposition of $G^\times$.

@@

	The main computational problem of this paper is \textsc{The Planar Minimum Branch Decomposition Problem}.

	\begin{definition}\label{def:mbdp}
		\textsc{The Planar Minimum Branch Decomposition Problem}

@@

		Output: A minimum branch decomposition of $G$.
	\end{definition}

	The algorithm described in this paper solves \textsc{The Planar Minimum Branch Decomposition Problem} in polynomial time.

@@

	Considering a plane graph $G$, you can compute a minimum branch decomposition $(B_G, \delta_G)$ of $G$ from a minimum carving decomposition $(C_{G^\times}, \lambda_{G^\times})$ of the medial graph $G^\times$ of $G$.

@@

	\begin{problem}\label{problem:medial}
		Given a plane graph $G$, output a medial graph $G^\times$, along with a bijectional relation between medial nodes $V(G^\times)$ and edges $\mathbb{E}(G)$.
	\end{problem}

	\begin{problem}\label{problem:mcd}
		Given a plane graph $M$, output a minimum carving decomposition of $M$.
	\end{problem}

	\begin{problem}\label{problem:mcd-to-mbd}
		Given a minimum carving decomposition of a medial graph of $G$, output a minimum branch decomposition of $G$.
	\end{problem}

@@

	\pyfile{branch_decomposition.py}{6}{22}

@@

		\pyfile{medial_graph.py}{5}{27}

	\subsection{Minimum Branch Decomposition from Minimum Carving Decomposition}

@@

	\subsection{Minimum Carving Decomposition}

@@

		By doing a series of edge contractions on a graph $M$, where the carving width does not increase until 3 vertices remain, then the series of contracted edges along with the three vertices, can be assembled into a minimum carving decomposition of $M$.

		The implementation finds a nonincreasing contraction by doing a linear search over every edge.

		No consideration has yet been given to any potential clever orderings of the edges that might improve the running time.

@@

		\pyfile{carving_decomposition.py}{7}{50}

@@

			\pyfile{carving_width.py}{7}{8}
			\pyfile{carving_width.py}{132}{158}

@@

		\pyfile{carving_width.py}{78}{130}

		\subsection{Quiet subgraph}

@@

		\subsection{Dual graph}\label{impl:dual}

@@

			The dual has a vertex for each face of the input graph. The faces are found by selecting an unmarked half-edge, then marking all the edges of the face it belongs to, and repeating this until all half-edges are marked.

@@

			\pyfile{dual_graph.py}{4}{56}

@@

	\todo{Agreement with Andreas' implementation on 6-100 vertices graphs, 10 different graphs}

	\todo{note Tested on cubic planar graph}

	\todo{tested by permuting vertex labels should not change the branch width}

	\todo{test that bw follows theoretical bound}

	Fomin and Thilikos 2006, New upper bound obn the decomposability pf planar graphs, bw <= 4.5 * sqrt(N) on planar

	Semour Thomas 1994, Call routing and ratcatcher, bw(G) = bw(dual(G)) assuming G is bridgeless planar graphs

	then the dual has half the vertices of the primal

	4.5 * sqrt(N/2)

	<=> 3.18199 * sqrt(n)

	Test bw <= 3.18199 * sqrt(N) on cubic planar graphs

	\todo{graphing theoretical bound}
